---
title: "Add Health Data Preparation for Within-Family Analysis of rGE"
author: "LiChen Dong"
date: "`r lubridate::today()`"
format: 
  html: 
    embed-resources: true
    toc: true 
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---


## Setup

```{r}
#| output: false # This tells R not to print the output of this code chunk

options(conflicts.policy = "depends.ok")
library(tidyverse)
library(lavaan)
theme_set(jtools::theme_apa())
path_data <- "graphing"
```

# Regression

### Read in data

```{r}
data <- readxl::read_xlsx(here::here(path_data, "coefficients_regression_EonG.xlsx"), trim_ws = F) |> 
        glimpse()

```

### Make `est` numeric and seperate `ci` into `lwr` and `upr`

```{r}
data <- data |> mutate(path = fct_rev(fct(path)),
                       text = est,
                       est = str_replace(est, "\\*$", ""),
                       lwr = str_split_i(ci, "\\s~\\s", 1),
                       upr = str_split_i(ci, "\\s~\\s", 2),
                       across(c(est,lwr,upr),as.numeric),
                       across(where(is.numeric), ~ round(., digits = 3)),
                       level = factor(level,c("within", "between", "population"),c("Within Family", "Between Family", "Population")))
```

### Forest Plot

```{r}
(rp1 <- ggplot(data, aes(y = est, ymin = lwr, ymax = upr, x = path, color = level, shape = level ,group = level)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(position = position_dodge(-.7), linewidth = .8, width = .3) +
  geom_point(position = position_dodge(-.7), size = 3, color = "black") +
  geom_text(aes(label = text, vjust = -.8), position = position_dodge(-.7), color = "black") +
  theme(axis.text.y = element_text(hjust=0)) + 
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.3, .3)) + 
  scale_color_brewer(palette = "Set1"))

ggsave("regression.png", rp1, path = here::here(path_data), width = 9.2, height = 7.8, dpi = 600)

```

### Forest Plot - Shorten

```{r}
(rp2 <- ggplot(data |> filter(path != "PGS → SUB"& path != "PGS → ASB"), aes(y = est, ymin = lwr, ymax = upr, x = path, color = level, shape = level ,group = level)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(position = position_dodge(-.7), linewidth = .8, width = .3) +
  geom_point(position = position_dodge(-.7), size = 3, color = "black") +
  geom_text(aes(label = text, vjust = -.8), position = position_dodge(-.7), color = "black") +
  theme(axis.text.y = element_text(hjust=0)) + 
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.3, .3)) + 
  scale_color_brewer(palette = "Set1"))

ggsave("regression_short.png", rp2, path = here::here(path_data), width = 9.2, height = 5.2, dpi = 600)

```

### Forest Plot - Population

```{r}
(rp3 <- ggplot(data |> filter(path != "PGS → SUB"& path != "PGS → ASB", level == "Population"), aes(y = est, ymin = lwr, ymax = upr, x = path, color = level, shape = level ,group = level)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(position = position_dodge(-.7), linewidth = .8, width = .15) +
  geom_point(position = position_dodge(-.7), size = 3, color = "black") +
  geom_text(aes(label = text, vjust = -.8), position = position_dodge(-.7), color = "black") +
  theme(axis.text.y = element_text(hjust=0)) + 
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.3, .3)) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(n = 3, name = "Set1")[3]))

ggsave("regression_pop.png", rp3, path = here::here(path_data), width = 9.2, height = 5.2, dpi = 600)

```

# Mediation_GPE



### Read in data
```{r}
data <- readxl::read_xlsx(here::here(path_data, "coefficients_mediation_GPE.xlsx"), trim_ws = F) |> 
        glimpse()

```


### Make `est` numeric and seperate `ci` into `lwr` and `upr`

```{r}
data <- data |> mutate(
                       model = paste(model, "Report"),
                       model = factor(model, 
                                      c("Parent Report", "Child Report")),
                       path = fct_rev(factor(path,
                                     c("Total PGS effect", 
                                       "Indirect effect via Wave 1 SUB",
                                       "Indirect effect via Wave 1 ASB",
                                       "Indirect effect via Childhood ADHD",
                                       "Indirect effects combined",
                                       "PGS direct effect"))),
                       text = est,
                       pct = replace_na(pct, ""),
                       est = str_replace(est, "\\*$", ""),
                       lwr = str_split_i(ci, "\\s~\\s", 1),
                       upr = str_split_i(ci, "\\s~\\s", 2),
                       text = paste0("b = ", text),
                       across(c(est,lwr,upr),as.numeric),
                       across(where(is.numeric), ~ round(., digits = 3)))
```

### Forest Plot

```{r}
(mp_gpe <- ggplot(data, aes(y = est, ymin = lwr, ymax = upr, x = path)) +
  geom_hline(yintercept = 0, color = "darkgrey", linewidth = .8, linetype = "dotted") +
  geom_errorbar(linewidth = .8, width = .1) +
  geom_point(size = 3, color = "black") +
  geom_text(data = data, aes(y = .1, label = text), vjust = -.8, hjust = 1, color = "black") +
  geom_text(data = data, aes(y = .1, label = pct), vjust = 1.2, hjust = 1, color = "black") +
  theme(axis.text.y = element_text(hjust=0)) + 
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.2, .1)) +
  facet_grid(. ~ model) + 
  scale_color_brewer(palette = "Set1"))

ggsave("mediation_gpe.png", mp_gpe, path = here::here(path_data), width = 12.2, height = 7.8, dpi = 600)

```









# Mediation_GEP



### Read in data
```{r}
data <- readxl::read_xlsx(here::here(path_data, "coefficients_mediation_GEP.xlsx"), trim_ws = F) |> 
        glimpse()

```


### Make `est` numeric and seperate `ci` into `lwr` and `upr`

```{r}
data <- data |> mutate(
                       model = factor(model, 
                                      c("Wave III SUB", "Wave III ASB")),
                       path = fct_rev(factor(path,
                                     c("Total PGS effect", 
                                       "Indirect effect via parent report",
                                       "Indirect effect via child report",
                                       "Indirect effects combined",
                                       "PGS direct effect"))),
                       text = est,
                       pct = replace_na(pct, ""),
                       est = str_replace(est, "\\*$", ""),
                       lwr = str_split_i(ci, "\\s~\\s", 1),
                       upr = str_split_i(ci, "\\s~\\s", 2),
                       text = paste0("b = ", text),
                       across(c(est,lwr,upr),as.numeric),
                       across(where(is.numeric), ~ round(., digits = 3)))
```

### Forest Plot

```{r}
(mp_gep <- ggplot(data, aes(y = est, ymin = lwr, ymax = upr, x = path)) +
  geom_hline(yintercept = 0, color = "darkgrey", linewidth = .8, linetype = "dotted") +
  geom_errorbar(linewidth = .8, width = .1) +
  geom_point(size = 3, color = "black") +
  geom_text(data = data, aes(label = text), vjust = -.8, color = "black") +
  geom_text(data = data, aes(label = pct), vjust = 1.8, color = "black") +
  theme(axis.text.y = element_text(hjust=0)) + 
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.15, .15)) +
  facet_grid(. ~ model) + 
  scale_color_brewer(palette = "Set1"))

ggsave("mediation_gep.png", mp_gep, path = here::here(path_data), width = 12.2, height = 7.8, dpi = 600)

```




















# end


## Read in data 

```{r}
data <- readxl::read_xlsx(here::here(path_data, "coefficients_mediation.xlsx"), trim_ws = F) |> 
        glimpse()
```

## Make `est` numeric and seperate `ci` into `lwr` and `upr`

```{r}
data <- data |> mutate(path = fct_rev(fct(path)),
                       est = str_replace(est, "\\*$", ""),
                       lwr = str_split_i(ci, "\\s~\\s", 1),
                       upr = str_split_i(ci, "\\s~\\s", 2),
                       across(c(est,lwr,upr),as.numeric),
                       across(where(is.numeric), ~ round(., digits = 3)),
                       level = factor(level,c("within", "between"),c("Within Family", "Between Family")))
```

## Forest Plot

```{r}
ggplot(data = data, aes(y = est, ymin = lwr, ymax = upr, x = path, color = level, shape = level ,group = level)) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_errorbar(position = position_dodge(.5), size = .8) +
  geom_point(position = position_dodge(.5), size = 2, color = "black") +
  theme(axis.text.y = element_text(hjust=0)) +
  facet_grid(. ~ model) +
  coord_flip(ylim = c(-.5, .5))
```



# Mediation GEP

### Read in data
```{r}
data <- readxl::read_xlsx(here::here(path_data, "coefficients_GEP.xlsx"), trim_ws = F) |> 
        glimpse()

```


### Make `est` numeric and seperate `ci` into `lwr` and `upr`

```{r}
data <- data |> mutate(model = paste("W3", model),
                       path = fct_rev(fct(path)),
                       text = est,
                       est = str_replace(est, "\\*$", ""),
                       lwr = str_split_i(ci, "\\s~\\s", 1),
                       upr = str_split_i(ci, "\\s~\\s", 2),
                       across(c(est,lwr,upr),as.numeric),
                       across(where(is.numeric), ~ round(., digits = 3)),
                       level = factor(level,c("within", "between", "population"),c("Within Family", "Between Family", "Population")))
```

### Forest Plot

```{r}
(mp_gep <- ggplot(data, aes(y = est, ymin = lwr, ymax = upr, x = path)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(linewidth = .8, width = .3) +
  geom_point(size = 3, color = "black") +
  geom_text(data = data |> filter(is.na(percent)), y = -0.1, aes(label = paste0("b = ", text)), hjust = 0, color = "black") +
  geom_text(data = data |> filter(!is.na(percent)), y = -0.1, aes(label = paste0("b = " , text, "\n", percent, " of total")), hjust = 0, color = "black") +
  theme(axis.text.y = element_text(hjust=0)) + 
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.1, .2)) +
  facet_grid(. ~ model) + 
  scale_color_brewer(palette = "Set1"))

ggsave("mediation_gep.png", mp_gep, path = here::here(path_data), width = 9.2, height = 5.2, dpi = 600)

```

# Mediation_GPE



### Read in data
```{r}
data <- readxl::read_xlsx(here::here(path_data, "coefficients_GPE.xlsx"), trim_ws = F) |> 
        glimpse()

```


### Make `est` numeric and seperate `ci` into `lwr` and `upr`

```{r}
data <- data |> mutate(
                       model = paste(model, "Report"),
                       model = factor(model, 
                                      c("Parent Report", "Child Report")),
                       path = fct_rev(fct(path)),
                       text = est,
                       est = str_replace(est, "\\*$", ""),
                       lwr = str_split_i(ci, "\\s~\\s", 1),
                       upr = str_split_i(ci, "\\s~\\s", 2),
                       across(c(est,lwr,upr),as.numeric),
                       across(where(is.numeric), ~ round(., digits = 3)),
                       level = factor(level,c("within", "between", "population"),c("Within Family", "Between Family", "Population")))
```

### Forest Plot

```{r}
(mp_gpe <- ggplot(data, aes(y = est, ymin = lwr, ymax = upr, x = path)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(linewidth = .8, width = .3) +
  geom_point(size = 3, color = "black") +
  geom_text(data = data |> filter(is.na(percent)), y = 0.1, aes(label = paste0("b = ", text)), hjust = 1, color = "black") +
  geom_text(data = data |> filter(!is.na(percent)), y = 0.1, aes(label = paste0("b = " , text, "\n", percent, " of total")), hjust = 1, color = "black") +
  theme(axis.text.y = element_text(hjust=0)) + 
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.2, .1)) +
  facet_grid(. ~ model) + 
  scale_color_brewer(palette = "Set1"))

ggsave("mediation_gpe.png", mp_gpe, path = here::here(path_data), width = 9.2, height = 5.2, dpi = 600)

```



## Read in data 

```{r}
data <- readxl::read_xlsx(here::here(path_data, "coefficients_mediation.xlsx"), trim_ws = F) |> 
        glimpse()
```

## Make `est` numeric and seperate `ci` into `lwr` and `upr`

```{r}
data <- data |> mutate(path = fct_rev(fct(path)),
                       est = str_replace(est, "\\*$", ""),
                       lwr = str_split_i(ci, "\\s~\\s", 1),
                       upr = str_split_i(ci, "\\s~\\s", 2),
                       across(c(est,lwr,upr),as.numeric),
                       across(where(is.numeric), ~ round(., digits = 3)),
                       level = factor(level,c("within", "between"),c("Within Family", "Between Family")))
```

## Forest Plot

```{r}
ggplot(data = data, aes(y = est, ymin = lwr, ymax = upr, x = path, color = level, shape = level ,group = level)) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_errorbar(position = position_dodge(.5), size = .8) +
  geom_point(position = position_dodge(.5), size = 2, color = "black") +
  theme(axis.text.y = element_text(hjust=0)) +
  facet_grid(. ~ model) +
  coord_flip(ylim = c(-.5, .5))
```


## PGS -> Env

```{r}
(p1 <- ggplot(data = data |> 
         filter(path %in% c("Total PGS effect on Parent-P",
                            "Total PGS effect on Parent-C",
                            "Total PGS effect on Peer",
                            "Total PGS effect on School"), model == "ASB"),
       aes(y = est, ymin = lwr, ymax = upr, x = path, color = level, shape = level ,group = level)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(position = position_dodge(-.5), linewidth = .8, width = .3) +
  geom_point(position = position_dodge(-.5), size = 3, color = "black") +
  theme(axis.text.y = element_text(hjust=0)) +
  labs(x = "Path", y = "Estimate") + 
  coord_flip(ylim = c(-.3, .3)))

(t1<- ggpubr::ggtexttable(data |> 
         filter(path %in% c("Total PGS effect on Parent-P",
                            "Total PGS effect on Parent-C",
                            "Total PGS effect on Peer",
                            "Total PGS effect on School") & 
                  model == "ASB" & level == "Within Family") |> 
         select(path, est, ci) , 
         rows = NULL, theme = ggpubr::ttheme("lBlueWhite")) |> 
      ggpubr::tab_add_title("Within Family Effects"))

(t2<- ggpubr::ggtexttable(data |> 
         filter(path %in% c("Total PGS effect on Parent-P",
                            "Total PGS effect on Parent-C",
                            "Total PGS effect on Peer",
                            "Total PGS effect on School") & 
                  model == "ASB" & level == "Between Family") |> 
         select(path, est, ci) , 
         rows = NULL, theme = ggpubr::ttheme("lBlueWhite")) |> 
      ggpubr::tab_add_title("Between Family Effects"))

(g1 <- ggpubr::ggarrange(p1, ggpubr::ggarrange(t1, t2, nrow = 2), ncol = 2, widths = c(2, 1)))

ggsave("totalpgs_env.png", g1, path = here::here(path_data), width = 12, height = 5, dpi = 600)
```


## PGS -> EXT I -> Env

```{r}
(p2 <- ggplot(data = data |> 
         filter(path %in% c("Total PGS effect on Parent-P",
                            "    PGS → W1 EXT → Parent-P",
                            "    PGS → Parent-P") &  level == "Within Family"), 
       aes(y = est, ymin = lwr, ymax = upr, x = path)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(position = position_dodge(-.5), linewidth = .8, width = .3) +
  geom_point(position = position_dodge(-.5), size = 3, color = "black") +
  theme(axis.text.y = element_text(hjust=0), plot.subtitle = element_text(hjust = 0.5)) +
  labs(x = "Path", y = "Estimate", subtitle = "Parenting - Parent Report") + 
  facet_grid(. ~ model) +
  coord_flip(ylim = c(-.2, .2)))

(p3 <- ggplot(data = data |> 
         filter(path %in% c("Total PGS effect on Parent-C",
                            "    PGS → W1 EXT → Parent-C",
                            "    PGS → Parent-C") &  level == "Within Family"), 
       aes(y = est, ymin = lwr, ymax = upr, x = path)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(position = position_dodge(-.5), linewidth = .8, width = .3) +
  geom_point(position = position_dodge(-.5), size = 3, color = "black") +
  theme(axis.text.y = element_text(hjust=0), plot.subtitle = element_text(hjust = 0.5)) +
  labs(x = "Path", y = "Estimate", subtitle = "Parenting - Child Report") + 
  facet_grid(. ~ model) +
  coord_flip(ylim = c(-.2, .2)))

(p4 <- ggpubr::ggarrange(p2, p3, ncol = 1))
ggsave("extmediate_pgs_env.png", p4, path = here::here(path_data), width = 8, height = 5, dpi = 600)

```

## PGS -> Env -> EXT III

```{r}
(p5 <- ggplot(data = data |> 
         filter(path %in% c("Total PGS effect on W3 EXT controlling for W1 EXT",
                            "    All indirect effects via Parent-P",
                            "    All indirect effects via Parent-C",
                            "    All indirect effects via Peer",
                            "    All indirect effects via School",
                            "    PGS → W3 EXT") &  level == "Within Family"), 
       aes(y = est, ymin = lwr, ymax = upr, x = path)) +
  geom_hline(yintercept = 0, color = "black", linewidth = .8, linetype = "dashed") +
  geom_errorbar(position = position_dodge(-.5), linewidth = .8, width = .3) +
  geom_point(position = position_dodge(-.5), size = 3, color = "black") +
  theme(axis.text.y = element_text(hjust=0), plot.subtitle = element_text(hjust = 0.5)) +
  labs(x = "Path", y = "Estimate", subtitle = "Parenting - Parent Report") + 
  facet_grid(. ~ model) +
  coord_flip(ylim = c(-.15, .15)))
```

